

\documentclass[11pt]{article}   
\usepackage{fullpage}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{xcolor}
\usepackage{float}
\restylefloat{table}

\newcommand{\F}{\mathbb{F}}
\newcommand{\np}{\mathop{\rm NP}}
\newcommand{\binom}[2]{{#1 \choose #2}}
\newcommand{\Z}{{\mathbb Z}}
\newcommand{\vol}{\mathop{\rm Vol}}
\newcommand{\conp}{\mathop{\rm co-NP}}
\newcommand{\atisp}{\mathop{\rm ATISP}}
\renewcommand{\vec}[1]{{\mathbf #1}}
\newcommand{\cupdot}{\mathbin{\mathaccent\cdot\cup}}
\newcommand{\mmod}[1]{\ (\mathrm{mod}\ #1)}  

\setlength{\parskip}{\medskipamount}
\setlength{\parindent}{0in}
%\input{dansmacs}


\begin{document}

\section*{CS 121 Homework Zero: Fall
2020}\label{cs-121-homework-zero-fall-2020}

The aim of problem set is to help you to test, and if needed to brush up, on the mathematical
background needed to be successful in CS 121. This problem set will be
due in the first week of class, but you are encouraged to start working on
it over the summer.

\textbf{Some policies:} (See the course policy at  {\tt http://madhu.seas.harvard.edu/courses/Fall2020/policy.html} for the
full policies.)

\begin{itemize}
\item
  {\bf Collaboration:} You can collaborate with other students that are currently enrolled in
  this course (or, in the case of homework zero, planning to enroll in
  this course) in brainstorming and thinking through approaches to
  solutions but you should write the solutions on your own and cannot
  share them with other students. 
\item
  {\bf Owning your solution:} Always make sure that you ``own'' your solutions to this other problem
  sets. That is, you should always first grapple with the problems on
  your own, and even if you participate in brainstorming sessions, make
  sure that you completely understand the ideas and details underlying
  the solution. This is in your interest as it ensures you have a solid
  understanding of the course material, and will help in the midterms
  and final. Getting 80\% of the problem
  set questions right on your own will be much better to both your
  understanding than getting 100\% of the questions through
  gathering hints from others without true understanding.
\item
  {\bf Serious violations:} Sharing questions or solutions with anyone outside this course,
  including posting on outside websites, is a violation of the honor
  code policy. Collaborating with anyone except students currently
  taking this course or using material from past years from this or
  other courses is a violation of the honor code policy.
\item
  {\bf Submission Format:} The submitted PDF should be typed and in the same format and
  pagination as ours. Please include the text of the problems and write
  \textbf{Solution X:} before your solution. Please mark in gradescope 
  the pages where
  the solution to each question appears. Points will be deducted if you
  submit in a different format.
\item {\bf Late Day Policy:} To give students some flexibility to manage your schedule, you are allowed a net total of {\bf eight} late days through the semester, but you may not take more than {\bf two} late days on any single problem set. No exceptions to this policy. 
\end{itemize}

\textbf{By writing my name here I affirm that I am aware of all policies
and abided by them while working on this problem set:}

\textbf{Your name:} Todd Morrill

\textbf{Collaborators:} (List here names of anyone you discussed
problems or ideas for solutions with)

% For homeworks 1 through n
\textbf{No. of late days used on previous psets (not including Homework Zero): }\\
\textbf{No. of late days used after including this pset: }


\newpage


\subsection*{Questions}\label{questions}

Please solve the following problems. Some of these might be harder than
the others, so don't despair if they require more time to think or you
can't do them all. Just do your best. Also, you should only attempt the
bonus questions if you have the time to do so. If you don't have a proof
for a certain statement, be upfront about it. You can always explain
clearly what you are able to prove and the point at which you were
stuck. Also, for a non bonus question, you can always simply write
\textbf{``I don't know''} and you will get 15 percent of the credit for
this problem.

The discussion board for this course will be active even before the course
starts. If you are stuck on this problem set, you can use this discussion board to send
a private message to all instructors under the \texttt{e-office-hours}
folder.

This problem set has a total of \textbf{50 points} and \textbf{11 bonus
points}. A grade of 50 or more on this problem set is considered a
perfect grade. If you get stuck in any questions, you might find the
resources in the CS 121 background page at {\tt https://cs121.boazbarak.org/background/} helpful.

\textbf{Problem 0 (5 points):} Read fully the \textbf{Mathematical
Background Chapter} (Chapter 1) from the textbook at 
{\tt http://madhu.seas.harvard.edu/courses/Fall2020/book.pdf}.
This is probably the most important exercise in this problem set!!

\textbf{Solution 0:} I certify that I fully read the
mathematical background chapter.

\newpage

\subsubsection{Logical operations, sets, and
functions}\label{logical-operations-sets-and-functions}

These questions assume familiarity with strings, functions, relations,
sets, and logical operators. We use an indexing from zero convention,
and so given a length \(n\) binary string \(x\in \{0,1\}^n\), we denote
coordinates of \(x\) by \(x_0,\ldots,x_{n-1}\). We use \([n]\) to denote
the set \(\{0,1,\ldots,n-1\}\).

\textbf{Question 1 (3 points):} Write a logical expression
\(\varphi(x)\) involving the variables \(x_0,x_1,x_2\) and the operators
\(\wedge\) (AND), \(\vee\) (OR), and \(\neg\) (NOT), such that
\(\varphi(x)\) is true if and only if the majority of the inputs are
\emph{False}.

\textbf{Solution 1:} $\varphi(x_0,x_1,x_2) = \neg(x_0 \wedge x_1) \wedge \neg(x_0 \wedge x_2) \wedge \neg(x_1 \wedge x_2)$

The proof is by truth table.
\begin{table}[H]
  \centering
  \begin{tabular}{|c|c|c|c|}
  \hline
  \multicolumn{1}{|l|}{$x_0$} &
    \multicolumn{1}{l|}{$x_1$} &
    \multicolumn{1}{l|}{$x_2$} &
    \multicolumn{1}{l|}{$\neg(x_0 \wedge x_1) \wedge \neg(x_0 \wedge x_2) \wedge \neg(x_1 \wedge x_2)$} \\ \hline
  T & T & T & F \\ \hline
  T & T & F & F \\ \hline
  T & F & T & F \\ \hline
  T & F & F & T \\ \hline
  F & T & T & F \\ \hline
  F & T & F & T \\ \hline
  F & F & T & T \\ \hline
  F & F & F & T \\ \hline
  \end{tabular}
  \end{table}

\newpage

\textbf{Question 2:} Use the logical quantifiers \(\forall\) (for all),
\(\exists\) (exists), as well as \(\wedge,\vee,\neg\) and the arithmetic
operations \(+,\times,=,>,<\) to write the following:

\textbf{Question 2.1 (3 points):} An expression \(\psi(n,k)\) such that
for every natural numbers \(n,k\), \(\psi(n,k)\) is true if and only if
\(k\) divides \(n\).

\textbf{Solution 2.1:}
$\psi(n,k) = \exists_{j \in \mathbb{N}}. n = j \cdot k$

\textbf{Question 2.2 (3 points bonus):} An expression \(\varphi(n)\)
such that for every natural number \(n\), \(\varphi(n)\) is true if and
only if \(n\) is a power of three.

\textbf{Solution 2.2:}
$\varphi(n) = \exists_{j \in \mathbb{N}}. \log_{3}(n) = j$

\newpage

\textbf{Question 3:} In this question, you need to describe in words
sets that are defined using a formula with quantifiers. For example, the
set \(S = \{ x\in \mathbb{N} \;:\; \exists_{y\in\mathbb{N}} x=2y \}\) is
the set of even numbers.

\textbf{Question 3.1 (3 points):} Describe in words the following set
\(S\):

\[
S = \{ x\in \{0,1\}^{100} : \forall_{i\in \{0,\ldots, 98\}} x_i = x_{i+1} \}
\]

(Recall that, as written in the mathematical background chapter, we use
zero-based indexing in this course, and so a string
\(x\in \{0,1\}^{100}\) is indexed as \(x_0x_1\cdots x_{99}\).)

\textbf{Solution 3.1:} $S$ is the set of strings of length 100 such that all letters
in the string are equal (i.e. all 0 or all 1).

\textbf{Question 3.2 (3 points):} Describe in words the following set
\(T\):

\[
T = \{ x\in \{0,1\}^* : |x|>1 \mbox{ and } \forall_{i \in \{2,\ldots,|x|-1 \} } \forall_{j \in \{2,\ldots,|x|-1\}} i\cdot j \neq |x| \}
\]

\textbf{Solution 3.2:} $T$ is the set of all bit strings of prime length.

\newpage

\textbf{Question 4:} This question deals with sets, their cardinalities,
and one to one and onto functions. You can cite results connecting these
notions from the course's textbook, MIT's ``Mathematics for Computer
Science'' or any other discrete mathematics textbook.

\textbf{Question 4.1 (4 points):} Define \(S = \{0,1\}^6\) and \(T\) as
the set \(\{ n \in [100] \;|\; \mbox{$n$ is prime } \}\). \emph{Prove or
disprove:} There is a one to one function from \(S\) to \(T\).

\textbf{Solution 4.1:}
$|S| = 2^6 = 64$ while $|T| = 25$. By the Pigeonhole Principle, there must be multiple 
elements $s \in S$ that map to some of the elements $t \in T$. Therefore, there is no
one-to-one function mapping $S$ to $T$.

\textbf{Question 4.2 (4 points):} Let \(n=100\),
\(S = [n] \times [n] \times [n]\) and \(T=\{0,1\}^n\). \emph{Prove or
disprove:} There is an onto function from \(T\) to \(S\).

\textbf{Solution 4.2:}
\begin{proof}
  $|T| = 2^{100}$ while $|S| = 10^6$. Since $|T| > |S|$, there is an onto function
  from $T$ to $S$.
\end{proof}

\textbf{Question 4.3 (4 points):} Let \(n=100\), let
\(S = \{0,1\}^{n^3}\) and \(T\) be the set of all functions mapping
\(\{0,1\}^n\) to \(\{0,1\}\). \emph{Prove or disprove:} There is a one
to one function from \(S\) to \(T\).

\textbf{Solution 4.3:}
\begin{proof}
  $|S| = 2^{100^3}$ while $|T| = 2^{2^{100}}$. Since $|S| < |T|$, there is an 
  one-to-one function from $S$ to $T$.
\end{proof}

\newpage

\textbf{Question 5.1 (5 points):} Prove that for every finite sets
\(A,B,C\), \(|A \cup B \cup C| \leq |A|+|B|+|C|\).

\textbf{Solution 5.1:}
\begin{proof}
  $|A \cup B \cup C| = |A|+|B|+|C| - |A\cap B| - |A\cap C| - |B\cap C| + |A\cap B \cap C|$ 
  by the Inclusion-Exclusion Principle. $|A\cap B|, |A\cap C|, |B\cap C| \geq 0$
  and $|A\cap B| + |A\cap C| + |B\cap C| \geq  |A\cap B \cap C|$ and therefore, 
  $|A \cup B \cup C| \leq |A|+|B|+|C|$, which is what we needed to show.
\end{proof}

\textbf{Question 5.2 (5 points bonus):} Prove that for every finite sets
\(A,B,C\),
\(|A \cup B \cup C| \geq |A|+|B|+|C| - |A \cap B| - |B \cap C| - |A \cap C|\).

\textbf{Solution 5.2:}
\begin{proof}
  $|A \cup B \cup C| = |A|+|B|+|C| - |A\cap B| - |A\cap C| - |B\cap C| + |A\cap B \cap C|$ 
  by the Inclusion-Exclusion Principle. $|A\cap B \cap C| \geq 0$ and therefore, 
  $|A \cup B \cup C| \geq |A|+|B|+|C| - |A \cap B| - |B \cap C| - |A \cap C|$, which is what we needed to show.
\end{proof}

\newpage

\subsubsection{Graphs}\label{graphs}

Thee following two questions assume familiarity with basic graph theory.
If you need to look up or review any terms, the CS 121 background page at
{\tt https://cs121.boazbarak.org/background/}
contains several freely available online resources on graph theory. This
material also appears in Chapters 13,14,16 and 17 of the CS 20 textbook
``Essential Discrete Mathematics for Computer Science'' by Harry Lewis
and Rachel Zax.

\textbf{Question 6.1 (5 points):} Prove that if \(G\) is a directed
acyclic graph (DAG) on \(n\) vertices, if \(u\) and \(v\) are two
vertices of \(G\) such that there is a directed path of length \(n-1\)
from \(u\) to \(v\) then \(u\) has no in-neighbors.\footnote{\emph{Hint:}
  You can use the topological sorting theorem shown in the mathematical
  background chapter.}

\textbf{Solution 6.1:}
\begin{proof}
  By Lemma 1.23 in the mathematical background chapter, if $G$ is acyclic then it
  has a layering. Let $f : V \rightarrow \mathbb{N}$ be a layering of $G$. 
  Suppose for the sake of contradiction that $u$ does have an in-neighbor, $w$, 
  which implies $f(w) < f(u)$. But $f(u) < f(w) < f(v)$ since there are there are 
  $n-1$ edges on the path from $u$ to $v$ and only $n$ vertices. But this is a 
  contradiction and therefore we can conlude that $u$ has no in-neighbors.
\end{proof}

\textbf{Question 6.2 (5 points):} Prove that for every undirected graph
\(G\) of \(1000\) vertices, if every vertex has degree at most \(4\),
then there exists a subset \(S\) of at least \(200\) vertices such that
no two vertices in \(S\) are neighbors of one another.

\textbf{Solution 6.2:}
\begin{proof}
  Suppose that we arbitrarily select vertices from $G$ to be removed
  and added to subset $S$. Each time we choose a vertex to be added to set $S$ we remove
  up to 5 vertices - the vertex to be added to set $S$ and the maximum of 4 vertices
  connected to it. This process can be repeated at least 200 times before exhausting
  all vertices in $G$.
\end{proof}

\newpage

\subsubsection{Big-O Notation}\label{big-o-notation}

\textbf{Question 7:} For each pair of functions \(f,g\) below, state
whether or not \(f=O(g)\) and whether or not \(g=O(f)\).

\textbf{Question 7.1 (3 points):} \(f(n)=n(\log n)^3\) and \(g(n)=n^2\).

\textbf{Solution 7.1:} $f = O(g)$, $g \neq O(f)$

\textbf{Question 7.2 (3 points):} \(f(n)= n^{\log n}\) and
\(g(n) = n^2\).

\textbf{Solution 7.2:} $g = O(f)$, $f \neq O(g)$

\textbf{Question 7.3 (3 points bonus):}
\(f(n) = \binom{n}{\lceil 0.2 n \rceil}\) (where \(\binom{n}{k}\) is the
number of \(k\)-sized subsets of a set of size \(n\)) and
\(g(n) = 2^{0.1 n}\).\footnote{\emph{Hint:} one way to do this is to use
  {Stirling's approximation} ({\tt https://en.wikipedia.org/wiki/Stirling\%27s\_approximation}).}

\textbf{Solution 7.3:} $g = O(f)$, $f \neq O(g)$

\end{document}


